Optimal Duplicate Detection in a Sliding Window

### 2020-03-27 qhtv2.tex
Various fixes, restructuration

\subsection{Related work}\label{sub:related}
The notion of sliding window was, as far as we know, first introduced in \cite{10.1145/1060745.1060753}. However, the definition is not the one we use, as several variations exist such as \cite{shtul2020agepartitioned}. Instead, the wDDP formulation we used was already introduced in \cite{Yoo10,She08}. Both of these papers propose an algorithm for the wDDP.

Closer to our work, the idea of using subfilters can be found in the A2 filter's design \cite{Yoo10} and a variation of our idea can be found in a very recent construction, on a slightly different DDP formulation \cite{shtul2020agepartitioned}. The A2 is built from two Bloom filters, a situation which we generalise and analyse generically in this paper. Similarly, the solution in \cite{shtul2020agepartitioned} only works with Bloom Filters.

\subsection{Organisation}
\todo

\section{Duplicate Detection and Filter Saturation}
\label{sec:dudefisa}
\subsection{Notations and Initial Problem Statement}
Let us consider an alphabet $\Gamma$. A stream $E$ over $\Gamma$ is a (possibly infinite) set $E = (e_1, e_2, \dotsc)$, with each $e_i \in \Gamma$.

@@

We usually consider the situations where the available memory is too small for perfect detection, i.e., $|\mathcal S| \ll |\Gamma|$. Otherwise, if $|\mathcal S| = |\Gamma|$ then solutions in constant time and without errors exist \cite{10.1145/3297280.3297335}.

@@

One can observe that $\FPR^w_n = \frac{1}{n}\sum_{i=1}^n \FP^w_i$, and similarly for $\FNR^w_n$.

@@

\subsection{DDP vs wDDP}
As noted before, some DDF were defined on the DDP, thus not offering a simple analysis of their error rates on the wDDP.

The sliding window variant of DDP may seem to be a slight weakening of the original problem. However, it is more appropriated for many use cases, such as caching or click fraud. % Remarkably, it turns out that DDFs which perform well for the DDP do rather poorly for the sliding window DDP.
A literature review collects the following DDF constructions: A2 filters \cite{Yoo10}, Stable Bloom Filters (SBF) \cite{Den06}, Quotient Hash Tables (QHT) \cite{10.1145/3297280.3297335}, Streaming Quotient Filters (SQF) \cite{Dut13}, Block-decaying Bloom Filters (b\_DBF) \cite{She08}, and a slight variation of Cuckoo Filters \cite{Fan14} suggested by \cite{10.1145/3297280.3297335}. The structure proposed in \cite{10.1145/1060745.1060753} is not desined for either wDDP or DDP, as they work on what they call `landmark` sliding window, which consists of a zero-resetting of the memory at some user-defined epochs.

Of these, only the A2 and b\_DBF were designed for the wDDP. The others do not account for a finite-length sliding window, and therefore cannot be adjusted as a function of $w$. Another remarkable (but experimental) observation is that these DDFs' false positive rate does not decrease to 0 when the sliding window becomes small.

@@

\subsection{Approximate Optimal wDDF}
In the previous section, the optimal filter requires to know the size of $\Gamma$, which might be unknown, unbounded or simply too big for practical applications. 
The dependence on $\log_2 |\Gamma|$ can be dropped, at the cost of allowing errors.

@@

	More precisely, it is possible to create a filter of $M$ bits with an FN of $0$, and an FP of $1 - (1-\frac1{w^2})^w \sim \frac 1w$, and a time complexity of $O(w)$.

	Using $M \simeq 5w\log_2 w$ bits of memory, a constant-time filter with the same error rate can be constructed.
\end{theorem}
Note that we only consider the probability of false positive after the filter has inserted at least $w$ elements, i.e., once the filter is full and has reached the stationary phase.

\begin{proof}
	Here again we explicitly construct the filters that attain the theorem's bounds.

	Let $h$ be a hash function with codomain $\{0,1\}^{2 \log_2 w}$.
The birthday theorem, as summed up in \cite{10.1007/3-540-45708-9_19}, states that for a hash function $h$ over $a$ bits, one must on average collect $2^{a/2}$ input-output pairs before obtaining a collision. Therefore $2^{(2 \log_2 w) / 2} = w$ hash values $h(e_i)$ can be computed before having a $50\%$ probability of a collision (here, a collision is when two distinct elements of the stream $e_i, e_j$ with $i \neq j, e_i \neq e_j$ have the same hash, i.e. $h(e_i) = h(e_j)$). The 50\% threshold we impose on $h$ is arbitrary but nonetheless practical.

@@

This gives us the Short Hash Filter (SHF), described in Algorithm~\ref{alg:SHF}.

\begin{algorithm}[]
	\caption{SHF Setup, Lookup and Insert}\label{alg:SHF}

@@

\subsection{Compact Short Hash Filter}
For the sake of completeness, we also introduce the Compact Short Hash Filter, which is very similar to the Short Hash Filter, but does not use any dictionnary. Hence, the filter is much slower (complexity of $O(w)$), but also has fewer errors since there is more place for the queue.

A quick description of the Compact Short Hash Filter (CSHF) is given in Algorithm~\ref{alg:CSHF}.

\begin{algorithm}[]
	\caption{CSHF, Lookup and Insert}\label{alg:CSHF}

@@

	\begin{algorithmic}[1]
		\Function{Insert}{$e$}
		\Comment Same as SHF, but removing all instructions about its dictionnary $D$.
		\EndFunction
	\end{algorithmic}
\end{algorithm}

\subsection{Error Rates}
\begin{theorem}
	Let $w \in \mathbb N$. Let $M \in \mathbb N$ be the available memory size. Let $\FN^w_{SHF}$ be the probability of false negative of an SHF with these parameters. We similarly define $\FP^w_{SHF}$, $\FN^w_{CSHF}$, $\FP^w_{CSHF}$. 
	Both SHF and QSHF have $\FN^w = 0$.

	Furthermore we have $\FP_{SHF}^w = 1 - \left(1 - \sqrt{w2^{-M/w}}\right)^w$ and $\FP_{CSHF}^w = 1 - \left(1 - 2^{-M/w}\right)^w$.
\end{theorem}

@@

It can be noted that a CSHF of size $M$ has the same error rate than an SHF of size $2M + w\log_2 w$.
\subsection{Saturation}
When benchmarking both structures, it appears they saturate very quickly when changing $w$: in other words, most SHF will either have an error rate of $0$ or $1$.

For instance, we benchmarked the performance of CSHF and SHF with a memory $M=100000$, on a stream of size 1 000 000 consisting of uniformly random elements from an alphabet of size $2^{18}$. Results are given in Figure~\ref{fig:shorts}.

\begin{figure}
	\input{graphs/bench_shorts.tex}
	\caption{Error rates of SHFs and CSHFs of $100000$ bits, for various $w$.}\label{fig:shorts}
\end{figure}

Finding an exact formula of the $w_{max}$ for which those filters reach an error of 50\% proved to be tedious. However, we numerically solved the equation $\FP^w = 0.5$ for these two filters, for about 200 different values of $M$, uniformly distributed \emph{on a log scale} between $10^2$ and $10^6$. Results are given in Figure~\ref{fig:w_opt}. As it turns out, the relation between $w_{max}$ and $M$ is quasi-linear, and we get the following linear regressions, with values rounded to the third significant digit:

\begin{figure}
	\input{graphs/w_opt_shf.tex}
	\caption{Value $w_{max}$ for which SHF and CSHF reach an error of 50\% ($\pm 0.001$), for various values of $M$.}\label{fig:w_opt}
\end{figure}

\begin{itemize}
	\item For CSHF, $w_{max}^{CSHF} = 0.0627 M + 443$, with an $r^2 = 0.9981$
	\item For SHF, $w_{max}^{SHF} = 0.0233 M + 186$, with an $r^2 = 0.9977$
\end{itemize}

\section{Queuing filters}\label{sec:queue}
In this section, we describe the queuing construction, which produces a sliding window DDF from any DDF. We first give the description of the setup, before studying the theoretical error rates. 
A scheme describing our structure is detailed in Figure~\ref{fig:scheme}.

@@

\subsection{Saturation Resistance of DDFs}\label{sub:saturation}
We evaluate the saturation rate for several DDFs, in the original DDP setting (without sliding window). Parameters are chosen to yield equivalent memory footprints and agree with \cite{10.1145/3297280.3297335}, namely:
%
\begin{itemize}
	\item QHT \cite{10.1145/3297280.3297335}, 1 bucket per row, 3 bits per fingerprint.
	\item SQF \cite{Dut13}, 1 bucket per row, $r = 2$ and $r' = 1$
	\item Cuckoo Filter \cite{Fan14}, cells containing 1 element of 3 bits each
	\item Stable Bloom Filter (SBF) \cite{Den06}, 2 bits per cell, 2 hash functions, targeted FPR of 0.02
	\item $A2$ Filter \cite{Yoo10}, targeted FPR of $0.1$ on the sliding window.
%	\item Block-Decaying Bloom Filter (b\_DBF) \cite{She08}, sliding window of 6000 elements.
\end{itemize}
%
These filters are run against a stream of uniformly sampled elements from an alphabet of $2^{26}$ elements. This results in around 8\% duplicates (up to statistical fluctuations) amongst the 150 000 000 elements in the longest stream used. %As a comparison, we also run the same benchmark against a stream obtained from crawling data, the same one as \cite{10.1145/3297280.3297335}, a real-world source which is highly non-uniform --- note that in this case, our lower bound on the error rate does not apply, owing to the stream being compressible. In that stream, around 10\% were duplicates amongst the 150 000 000 elements.
Results are plotted in Figure~\ref{fig:graph_n}.

\begin{figure}
	\begin{tikzpicture}
	\input{er_n_artificial}
	\end{tikzpicture}
	\caption{Error rate (times 100) of DDFs of 1Mb, over an artificial stream, depending on the size of the stream, without any sliding window. Hatched area represents over-optimal (impossible) values.}\label{fig:graph_n}
\end{figure}

>>>>

\subsection{Related work}\label{sub:related}
The notion of sliding window was, as far as we know, first introduced in \cite{10.1145/1060745.1060753}. However, several variations exist and describe incompatible situations, such as \cite{shtul2020agepartitioned}. Instead, the wDDP formulation we rely on is due to \cite{Yoo10,She08}, which also introduce algorithms for solving the wDDP approximately.

The notion of using subfilters, as in the queuing construction, can be found in the A2 filter's design \cite{Yoo10} and a variation thereof can be found in \cite{shtul2020agepartitioned} but in a different DDP formulation. The A2 is built from two Bloom filters, a situation which we generalise and analyse generically in this paper. Similarly, the construction in \cite{shtul2020agepartitioned} only works with Bloom Filters.

\section{Duplicate Detection and Filter Saturation}
\label{sec:dudefisa}
\subsection{Notations and basic definitions}
Let us consider an alphabet $\Gamma$. A stream $E$ over $\Gamma$ is a (possibly infinite) set $E = (e_1, e_2, \dotsc)$, with each $e_i \in \Gamma$.

@@

We usually consider the situations where the available memory is too small for perfect detection, i.e., $|\mathcal S| \ll |\Gamma|$. Otherwise, if $|\mathcal S| = |\Gamma|$ then the problem can be solved in constant time without errors \cite{10.1145/3297280.3297335}.

@@

We have $\FPR^w_n = \frac{1}{n}\sum_{i=1}^n \FP^w_i$, and similarly for $\FNR^w_n$.

@@

\section{DDFs and the wDDP}
A literature review collects the following DDF constructions: A2 filters \cite{Yoo10}, Stable Bloom Filters (SBF) \cite{Den06}, Quotient Hash Tables (QHT) \cite{10.1145/3297280.3297335}, Streaming Quotient Filters (SQF) \cite{Dut13}, Block-decaying Bloom Filters (b\_DBF) \cite{She08}, and a slight variation of Cuckoo Filters \cite{Fan14} suggested by \cite{10.1145/3297280.3297335}. The structure proposed in \cite{10.1145/1060745.1060753} is not desined for wDDP but a variant called `landmark` sliding window, which consists of a zero-resetting of the memory at some user-defined epochs.

Of these, only the A2 and b\_DBF were designed explicitly for the wDDP. Others DDFs do not account for a finite-length sliding window, and therefore cannot be adjusted as a function of $w$. Another remarkable (but experimental) observation is that these DDFs' false positive rate does not decrease to 0 when the sliding window becomes small.

@@

\subsection{Approximate Optimal wDDF}
The optimal filter described above requires that the size of $\Gamma$ is known in advance. The dependence on $\log_2 |\Gamma|$ can be dropped, at the cost of allowing errors.

@@

	More precisely, it is possible to create a filter of $M$ bits with an FN of $0$, an FP of $1 - (1-\frac1{w^2})^w \sim \frac 1w$, and a time complexity of $O(w)$.

	Using $M \simeq 5w\log_2 w$ bits of memory, a constant-time filter with the same error rate can be constructed.
\end{theorem}
Note that we only consider the probability of false positive after the filter has inserted at least $w$ elements, i.e., once the filter is full and has reached a stationary regime.
\begin{proof}
	Here again we explicitly construct the filters that attain the theorem's bounds.

	Let $h$ be a hash function with codomain $\{0,1\}^{2 \log_2 w}$.
The birthday theorem \cite{10.1007/3-540-45708-9_19} states that for a hash function $h$ over $a$ bits, one must on average collect $2^{a/2}$ input-output pairs before obtaining a collision. Therefore $2^{(2 \log_2 w) / 2} = w$ hash values $h(e_i)$ can be computed before having a $50\%$ probability of a collision (here, a collision is when two distinct elements of the stream $e_i, e_j$ with $i \neq j, e_i \neq e_j$ have the same hash, i.e. $h(e_i) = h(e_j)$). The 50\% threshold we impose on $h$ is arbitrary but nonetheless practical.

@@

This gives us the Short Hash Filter (SHF), described in Algorithm~\ref{alg:SHF}. The implementation relies on a double-ended queue or a ring buffer, which allows pushing at beggining of a queue and popping at the end in constant time.

\begin{algorithm}[!ht]
	\caption{SHF Setup, Lookup and Insert}\label{alg:SHF}

@@

\subsection{Compact Short Hash Filter}
Removing the dictionary from the SHF construction yields a more memory-efficient, but less time-efficient construction, which we dub \enquote{compact} short hash filter (CSHF). The CSHF performs in linear time in $w$ and is described in Algorithm~\ref{alg:CSHF}.

\begin{algorithm}[!ht]
	\caption{CSHF, Lookup and Insert}\label{alg:CSHF}

@@

	\begin{algorithmic}[1]
		\Function{Insert}{$e$}
		\State $Q.\mathsf{Push\_Front}(h(e))$

		\If{$Q.\mathsf{length}() > w$}
			\State $h' \gets Q.\mathsf{Pop\_back}()$
		\EndIf
		\EndFunction
	\end{algorithmic}
\end{algorithm}

\subsection{Error Rates}
Let $w> 0$ be a window size and $M > 0$ the available memory.

We write $\FN^w_\text{SHF}$ the probability of false negative of an SHF with these parameters. We similarly define $\FP^w_\text{SHF}$, $\FN^w_\text{CSHF}$, $\FP^w_\text{CSHF}$.

\begin{theorem}	
	\begin{align*}
		\FN_\text{SHF}^w&  = \FN_\text{CSHF}^w =0 \\
		\FP_\text{SHF}^w & = 1 - \left(1 - \sqrt{w2^{-M/w}}\right)^w \\
		\FP_\text{CSHF}^w & = 1 - \left(1 - 2^{-M/w}\right)^w
	\end{align*}
\end{theorem}

@@

\paragraph{Remark:} A CSHF of size $M$ has the same error rate than an SHF of size $2M + w\log_2 w$.
\subsection{Saturation}
Both SHF and CSHF have strictly increasing error rates, which reach a threshold of $1/2$ for some maximum window size $w_\text{max}$. Beyond this value, these filters saturate extremely quickly: in other words, most SHF will either have an error rate of $0$ or $1$.

An illustration of this phenomenon can be seen in Figure~\ref{fig:shorts}, 
which shows the error rates for CSHF and SHF with $M=100000$, against 
a uniformly random stream of $18$-bit elements ($|\Gamma| = 2^{18}$). The benchmark used a finite stream of length $10^6$.

\begin{figure}[!ht]
	\centering
	\input{graphs/bench_shorts.tex}
	\caption{Error rates of SHFs and CSHFs for $M = 10^5$ bits, for various window sizes $w$.}\label{fig:shorts}
\end{figure}

The value $w_\text{max}$ can be obtained by solving (numerically) for $\FP^{w_\text{max}} = 1/2$ for a given $M$. Experimentally, we observed an approximately linear relationship between $M$ and $w_\text{max}$ for large values of $M$.

\section{Non-windowed DDFs in a wDDP setting}
\subsection{Saturation Resistance of DDFs}\label{sub:saturation}
We evaluate the saturation rate for several DDFs, in the original DDP setting (without sliding window). Parameters are chosen to yield equivalent memory footprints and agree with \cite{10.1145/3297280.3297335}, namely:
%
\begin{itemize}
	\item QHT \cite{10.1145/3297280.3297335}, 1 bucket per row, 3 bits per fingerprint.
	\item SQF \cite{Dut13}, 1 bucket per row, $r = 2$ and $r' = 1$
	\item Cuckoo Filter \cite{Fan14}, cells containing 1 element of 3 bits each
	\item Stable Bloom Filter (SBF) \cite{Den06}, 2 bits per cell, 2 hash functions, targeted FPR of 0.02
	\item $A2$ Filter \cite{Yoo10}, targeted FPR of $0.1$ on the sliding window.
%	\item Block-Decaying Bloom Filter (b\_DBF) \cite{She08}, sliding window of 6000 elements.
\end{itemize}
%
These filters are run against a stream of uniformly sampled elements from an alphabet of $2^{26}$ elements. This results in around 8\% duplicates (up to statistical fluctuations) amongst the 150 000 000 elements in the longest stream used. %As a comparison, we also run the same benchmark against a stream obtained from crawling data, the same one as \cite{10.1145/3297280.3297335}, a real-world source which is highly non-uniform --- note that in this case, our lower bound on the error rate does not apply, owing to the stream being compressible. In that stream, around 10\% were duplicates amongst the 150 000 000 elements.
Results are plotted in Figure~\ref{fig:graph_n}.

\begin{figure}[!ht]
	\begin{tikzpicture}
	\input{er_n_artificial}
	\end{tikzpicture}
	\caption{Error rate (times 100) of DDFs of 1Mb as a function of stream length. Hatched area represents over-optimal (impossible) values.}\label{fig:graph_n}
\end{figure}

\subsection{Performance in wDDF}
We now consider the performance of the filters just discussed in the \emph{windowed} setting, for which they were \emph{not} designed. In particular, it is not possible to adjust their parameters as a function of $w$.

Remarkably, some of these filters still outperform dedicated windowed filters for some window sizes at least, as shown in Figure~\label{fig:improvement}. Nevertheless, we will now discuss the queuing construction, which allow us to build windowed filters from these.

\section{Queuing filters}\label{sec:queue}
We now describe the queuing construction, which produces a sliding window DDF from any DDF. We first give the description of the setup, before studying the theoretical error rates. 
A scheme describing our structure is detailed in Figure~\ref{fig:scheme}.

@@
